Arrows for Parallel Computations

### 2017-07-11 paper/content/topologySkeletons.tex
wip on the paper. troubles with Par monad are gone

@@

The parallel |pipe| skeleton is semantically equivalent to folding over a list |[arr a a]| of arrows with |>>>|, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine. We implement this skeleton using the |ArrowLoop| typeclass which gives us the |loop :: arr (a, b) (c, b) -> arr a c| combinator which allows us to express recursive fix-point computations in which output values are fed back as input. For example %this
\begin{code}
loop (arr (\(a, b) -> (b, a:b)))
\end{code}
which is the same as
\begin{code}
loop (arr snd &&& arr (uncurry (:)))
\end{code}
defines an arrow that takes its input |a| and converts it into an infinite stream |[a]| of it. Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |parEvalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.

@@

\begin{figure}[t]
\begin{code}
pipeSimple :: (ArrowLoop arr, ArrowParallel arr a a conf) =>
	conf -> [arr a a] -> arr a a
pipeSimple conf fs =
	loop (arr snd &&&
		(arr (uncurry (:) >>> lazy) >>> parEvalN conf fs)) >>>
	arr last
\end{code}
\caption{A~first implementation of the |pipe| skeleton expressed with Parallel Arrows. Note that the use of |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) is essential as without it programs using this definition would never halt. We need to enforce that the evaluation of the input |[a]| terminates before passing it into |parEvalN|.}
\label{fig:pipeSimple}

\begin{code}
pipe :: (ArrowLoop arr, ArrowParallel arr (fut a) (fut a) conf, Future fut a) =>
	conf -> [arr a a] -> arr a a
pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
\end{code}
\caption{Final definition of the |pipe| skeleton with Futures.}
\label{fig:pipe}
\end{figure}

Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e. |arr a b| and |arr b c|. By wrapping these two arrows inside a common type we obtain |pipe2| (Fig.~\ref{fig:pipe2}).
\begin{figure}[tb]
\begin{code}
pipe2 :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
         ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf) =>
	conf -> arr a b -> arr b c -> arr a c
pipe2 conf f g =
	(arr return &&& arr (const [])) &&& arr (const []) >>>
	pipe conf (replicate 2 (unify f g)) >>>
	arr snd >>> arr head where
			unify :: (ArrowChoice arr) => arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
			unify f g =
				(mapArr f *** mapArr g) *** arr (\_ -> []) >>>
				arr (\((a, b), c) -> ((c, a), b))

(parcomp) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) ()) =>
	arr a b -> arr b c -> arr a c
(parcomp) = pipe2 ()
\end{code}
\caption{Definition of |pipe2| and a parallel |>>>|.}
\label{fig:pipe2}
\end{figure}

Note that extensive use of |pipe2| over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN. Nonetheless, we can define a version of parallel piping operator |parcomp|, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.

@@

Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into a our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| before lifting it to |arr [(i, fut r)] [(o, fut r)]| to get a list |[(o, fut r)]|. Finally we unzip this list into |([o], [fut r])|. Plugging this arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring arrow (Fig.~\ref{fig:ringFinal}).
This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.

Note that the Par Monad version of the |ArrowParallel| version  presented in this paper (Fig.~\ref{fig:ArrowParallelParMonad}) does not work with the looping behaviour in the |ring| skeletons. Whenever |runPar| is invoked, strictness is enforced, which means that the |lazy| operation in the definition of |ring| will not work.
%Further details on this can be found in \cite{eden_cefp}.

\begin{figure}[tb]
\begin{code}
ring :: (ArrowLoop arr, Future fut r, ArrowParallel arr (i, fut r) (o, fut r) conf) =>
    conf -> arr (i, r) (o, r) -> arr [i] [o]
ring conf f =
	loop (second (rightRotate >>> lazy) >>> arr (uncurry zip) >>>
        parMap conf (second get >>> f >>> second put) >>> arr unzip)
\end{code}
\caption{Final definition of the |ring| skeleton.}
\label{fig:ringFinal}
\end{figure}
 % $ %% formatting
% and lazy:
% \begin{code}
% lazy :: (Arrow arr) => arr [a] [a]
% lazy = arr (\ ~(x:xs) -> x : lazy xs
% \end{code}
%% we have shown this already.

@@

This action is, however,  more complicated than in the ring case as we have one more dimension of inputs to be transformed. We first have to |shuffle| all the inputs to then pass it into |parMap conf (ptorus f)| which yields |[(d, fut a, fut b)]|. We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| typeclass. Finally, we unpack this matrix |[[[(d, fut a, fut b)]]| with |arr (map unzip3) >>> arr unzip3 >>> threetotwo| to get  |([[d]], ([[fut a]], [[fut b]]))|.

\begin{figure}[tb]
\begin{code}
torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr, Future fut a, Future fut b,
          ArrowParallel arr (c, fut a, fut b) (d, fut a, fut b) conf) =>
	conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
torus conf f =
	loop (second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy)) >>>
	arr (uncurry3 (zipWith3 lazyzip3)) >>>
	(arr length >>> arr unshuffle) &&& (shuffle >>> parMap conf (ptorus f)) >>> app >>>
	arr (map unzip3) >>> arr unzip3 >>> threetotwo)

@@

Again, the implementation of the Par Monad version prevents this definition of the |torus| skeleton to work due to the reasons explained in the section about the |ring| skeleton in section \ref{sec:ring}.

>>>>

If we used the original definition of |parEvalN|, however, these skeletons would produce an infinite loop with the Multicore and Par Monad backends which during runtime would result in the program crashing. This materializes with the usage of |loop| of the |ArrowLoop| typeclass and is probably due to the way their respective parallelism engines work internally. \mbcomment{okay so?} As these skeletons probably do not make any practical sense besides for testing with these backends anyways (because of the shared memory between the threads), we create an extra abstraction layer for the original |parEvalN| in these skeletons called |evalN| in the |FutureEval| typeclass. This allows us for selective enabling and disabling of parallelism.
\begin{code}
class ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN :: (ArrowParallel arr a b conf) => conf -> [arr a b] -> arr [a] [b]
\end{code}
As Eden has no problems with the looping skeletons, we declare a delegating instance:
\begin{code}
instance ArrowParallel arr a b conf => FutureEval arr a b conf where
    evalN = parEvalN
\end{code}
The Par Monad and Multicore backends have parallelism disabled in their instance of |FutureEval|. This way the skeletons can still run without errors on shared-memory machines and still be used to test programs locally.
\begin{code}
instance (Arrow arr, ArrowChoice arr, ArrowApply arr,
    ArrowParallel arr a b conf) => FutureEval arr a b conf where
    evalN _ = listApp
\end{code}

@@

The parallel |pipe| skeleton is semantically equivalent to folding over a list |[arr a a]| of arrows with |>>>|, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine. We implement this skeleton using the |ArrowLoop| typeclass which gives us the |loop :: arr (a, b) (c, b) -> arr a c| combinator which allows us to express recursive fix-point computations in which output values are fed back as input. For example %this
\begin{code}
loop (arr (\(a, b) -> (b, a:b)))
\end{code}
which is the same as
\begin{code}
loop (arr snd &&& arr (uncurry (:)))
\end{code}
defines an arrow that takes its input |a| and converts it into an infinite stream |[a]| of it. Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |evalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.

@@

\begin{figure}[t]
\begin{code}
pipeSimple :: (ArrowLoop arr, FutureEval arr a a conf) =>
	conf -> [arr a a] -> arr a a
pipeSimple conf fs =
	loop (arr snd &&&
		(arr (uncurry (:) >>> lazy) >>> evalN conf fs)) >>>
	arr last
\end{code}
\caption{A~first implementation of the |pipe| skeleton expressed with Parallel Arrows. Note that the use of |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) is essential as without it programs using this definition would never halt. We need to enforce that the evaluation of the input |[a]| terminates before passing it into |evalN|.}
\label{fig:pipeSimple}

\begin{code}
pipe :: (ArrowLoop arr, FutureEval arr (fut a) (fut a) conf, Future fut a) =>
	conf -> [arr a a] -> arr a a
pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
\end{code}
\caption{Final definition of the |pipe| skeleton with Futures.}
\label{fig:pipe}
\end{figure}

Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e. |arr a b| and |arr b c|. By wrapping these two arrows inside a common type we obtain |pipe2| (Fig.~\ref{fig:pipe2}).
\begin{figure}[tb]
\begin{code}
pipe2 :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
         FutureEval arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf) =>
	conf -> arr a b -> arr b c -> arr a c
pipe2 conf f g =
	(arr return &&& arr (const [])) &&& arr (const []) >>>
	pipe conf (replicate 2 (unify f g)) >>>
	arr snd >>> arr head where
			unify :: (ArrowChoice arr) => arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
			unify f g =
				(mapArr f *** mapArr g) *** arr (\_ -> []) >>>
				arr (\((a, b), c) -> ((c, a), b))

(parcomp) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
	FutureEval arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) ()) =>
	arr a b -> arr b c -> arr a c
(parcomp) = pipe2 ()
\end{code}
\caption{Definition of |pipe2| and a parallel |>>>|.}
\label{fig:pipe2}
\end{figure}

Note that extensive use of |pipe2| over |pipe| with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to |parEvalN| inside of |evalN|. Nonetheless, we can define a version of parallel piping operator |parcomp|, which is semantically equivalent to |>>>| similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.

@@

Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into a our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| before lifting it to |arr [(i, fut r)] [(o, fut r)]| to get a list |[(o, fut r)]|. Finally we unzip this list into |([o], [fut r])|. Plugging this arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring arrow (Fig.~\ref{fig:ringFinal}).
This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
%Further details on this can be found in \cite{eden_cefp}.

\begin{figure}[tb]
\begin{code}
ring :: (ArrowLoop arr, Future fut r, FutureEval arr (i, fut r) (o, fut r) conf) =>
    conf -> arr (i, r) (o, r) -> arr [i] [o]
ring conf f =
	loop (second (rightRotate >>> lazy) >>> arr (uncurry zip) >>>
        evalN conf (repeat (second get >>> f >>> second put)) >>> arr unzip)
\end{code}
\caption{Final definition of the |ring| skeleton.}
\label{fig:ringFinal}
\end{figure}
 % $ %% formatting
% and lazy:
% \begin{code}
% lazy :: (Arrow arr) => arr [a] [a]
% lazy = arr (\ ~(x:xs) -> x : lazy xs
% \end{code}
%% we have shown this already.

@@

This action is, however,  more complicated than in the ring case as we have one more dimension of inputs to be transformed. We first have to |shuffle| all the inputs to then pass it into |evalN conf (repeat (ptorus f))| which yields |[(d, fut a, fut b)]|. We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with |arr length >>> arr unshuffle| with the use of |app :: arr (arr a b, a) c| from the |ArrowApply| typeclass. Finally, we unpack this matrix |[[[(d, fut a, fut b)]]| with |arr (map unzip3) >>> arr unzip3 >>> threetotwo| to get  |([[d]], ([[fut a]], [[fut b]]))|.

\begin{figure}[tb]
\begin{code}
torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr, Future fut a, Future fut b,
          FutureEval arr (c, fut a, fut b) (d, fut a, fut b) conf) =>
	conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
torus conf f =
	loop (second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy)) >>>
	arr (uncurry3 (zipWith3 lazyzip3)) >>>
	(arr length >>> arr unshuffle) &&& (shuffle >>> evalN conf (repeat (ptorus f)) >>> app >>>
	arr (map unzip3) >>> arr unzip3 >>> threetotwo)

@@
